---
title: "Child Count and Census"
author: "Maiko Hata"
format: 
  pdf:
    mainfont: "Times New Roman"
    sansfont: "Times New Roman"
fig-width: 6
fig-height: 4
csl: apa.csl
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(tidyr)
library(readr)
```

1.  Import the excel sheet of racial groups represented in census 2020 and EI childcount 19-20

```{r}
child_count_vs_census <- import("/Users/hata/Desktop/EDLD652_Diss/Data/use this racial group representation in general population and in EI NATIONAL.xlsx", sheet = 2)
```

```{r}
print(child_count_vs_census)
```

```{r}
# colnames(child_count_vs_census)
```

```{r}
# I ACCIDENTALLY DELETED THE "-" IN cols = -`...1`!!! Good job, me, figuring it out with ChatGPT!!
child_count_vs_census_long <- child_count_vs_census %>% 
  pivot_longer(
    cols = -`...1`, 
    names_to = "Race", 
    values_to = "Percentage"
  )
```

```{r}
ggplot(child_count_vs_census_long, 
       aes(x = Race, 
           y = Percentage, 
           fill = `...1`)) +
  geom_bar(stat = "identity") +
  labs(title = "Child Count vs Census by Race",
       x = "Race",
       y = "Percentage",
       fill = "Location") +  
  theme_minimal()
```

```{r}
# stacked bar - added "dodge" 

library(ggplot2)

ggplot(child_count_vs_census_long, 
       aes(x = Race, 
           y = Percentage, 
           fill = `...1`)) +  
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Child Count vs Census by Race",
       x = "Race",
       y = "Percentage",
       fill = "Location") +
  theme_minimal()
```

```{r}
colnames(child_count_vs_census)
```

```{r}
# realphabetizing the X axis 
child_count_vs_census_long <- child_count_vs_census_long %>%
  mutate(Race = factor(Race, levels = c(
    "American Indian or Alaska Native", 
    "Asian", 
    "Black or African American", 
    "Hispanic/Latino", 
    "Native Hawaiian or Pacific Islander", 
    "Two or More Races", 
    "White"))) 
```

```{r}
child_count_vs_census_long <- child_count_vs_census_long %>%
  mutate(Race = fct_rev(Race))
```

```{r}
library(readr)

write_csv(child_count_vs_census_long, "/Users/hata/Desktop/EDLD652_Diss/Data/child_count_vs_census_long.csv")
```

```{r}
# flipping x and y 
ggplot(child_count_vs_census_long, 
       aes(x = Percentage, 
           y = Race, 
           fill = `...1`)) +  
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Child Count vs Census by Race",
       x = "Percentage",
       y = "Race",
       fill = "Location") +
  theme_minimal(base_family = "Times New Roman")
```

```{r}
library(ggplot2)
library(viridis)

ggplot(child_count_vs_census_long, 
       aes(x = Percentage, 
           y = Race, 
           fill = `...1`)) +  
  geom_bar(stat = "identity", position = "dodge") +  
  geom_text(aes(label = Percentage),  # Add data labels
            position = position_dodge(width = 0.9),  # Align labels with bars
            vjust = -0.3,  # Position labels above bars
            size = 3,       # Adjust text size if needed
            family = "Times New Roman") +  # Font for labels
  
  labs(title = "Child Count vs Census by Race",
       x = "Percentage",
       y = "Race",
       fill = "Location") +
  theme_minimal(base_family = "Times New Roman")

```

```{r}
ggplot(child_count_vs_census_long, 
       aes(x = Percentage, 
           y = Race, 
           fill = `...1`)) +  
  geom_bar(stat = "identity", position = "dodge") +  
  geom_text(aes(label = Percentage),  
            position = position_dodge(width = 0.9),  
            hjust = -0.2,  # Move labels slightly to the right of bars
            size = 3,       
            family = "Times New Roman") +  

  labs(title = "Child Count vs Census by Race",
       x = "Percentage",
       y = "Race",
       fill = "Location") +
  theme_minimal(base_family = "Times New Roman")
```

## Child count vs census (2020) BUT BY NUMBERS not % so I can run chi-square

```{r}
child_count_vs_census_number <- import("/Users/hata/Desktop/EDLD652_Diss/Data/use this racial group representation in general population and in EI NATIONAL.xlsx", sheet = 3)
```

```{r}
head(child_count_vs_census_number)
```

```{r}
write.csv(child_count_vs_census_number, 
          file = "/Users/hata/Desktop/EDLD652_Diss/Data/childcount_vs_census_number.csv", row.names = FALSE)
```

### Making child_count_vs_census_number into matrix so I can run corrplot().

-   Save it as matrix

-   Drop the total, not needed for chi-square

```{r}
child_count_vs_census_number <- child_count_vs_census_number %>% 
  select(-2)
```

-   Set the row names from first column, because for a chi-square test, you want this stored as a **matrix with named rows**. So, rownames(df_chi) \<- df_chi\[\[1\]\] Takes the first column (e.g., `c("General Population", "EI Population")`) and assigns those as **row names** of the data frame to prepare the structure for conversion to a matrix.

```{r}
rownames(child_count_vs_census_number) <- child_count_vs_census_number[[1]]
```

-   Drop the first column now that it's row names

```{r}
child_count_vs_census_number <- child_count_vs_census_number %>% 
  select(-1)
```

-   Convert to a matrix

```{r}
child_count_vs_census_number_matrix <- as.matrix(child_count_vs_census_number)
```

-   Run chi-square and see the results

```{r}
chi_results_cc_vs_census_number <- chisq.test(child_count_vs_census_number_matrix)

chi_results_cc_vs_census_number
```

Pearson's Chi-squared test

data: child_count_vs_census_number_matrix X-squared = 26956, df = 6, p-value \< 2.2e-16

-   View standardized residues

```{r}
chi_results_cc_vs_census_number$stdres
```

-   Save the results with standardized residues

```{r}
chisq.test(child_count_vs_census_number_matrix, 
           correct = FALSE)$stdres
```

```{r}
adj_std_res_race_us_representation <- 
  chisq.test(child_count_vs_census_number_matrix)
```

```{r}
plot.new()
dev.off()
```

```{r}
print(chi_results_cc_vs_census_number$stdres)
```

-   The below didn't run first because it had "/" in "Hispanic/Latino" so I changed it to "... or ..."

```{r}
colnames(chi_results_cc_vs_census_number$stdres) <- c(
  "American Indian or Alaska Native", 
"Asian", 
  "Black or African American", 
  "Hispanic or Latino", 
  "Native Hawaiian or Pacific Islander", 
  "Two or More Races", 
  "White")
```

```{r}
rownames(chi_results_cc_vs_census_number$stdres) <- c(
  "General Population", 
  "EI Population"
)
```

```{r}
chi_results_cc_vs_census_number$stdres <- chi_results_cc_vs_census_number$stdres[, order(colnames(chi_results_cc_vs_census_number$stdres))]
```

```{r}
corrplot(chi_results_cc_vs_census_number$stdres, 
         is.cor = FALSE, 
         tl.cex = 0.8, 
         tl.col = "black", 
         cl.cex = 0.7,     
         cl.offset = 1,    
         cl.ratio = 0.2)  
```

```{r}
corrplot(
  t(chi_results_cc_vs_census_number$stdres),  # transpose the matrix
  is.cor = FALSE, 
  tl.cex = 0.8, 
  tl.col = "black", 
  cl.cex = 0.7,     
  cl.offset = 1,    
  cl.ratio = 0.2
)

```

## Odds ratio for representation by race

```{r}

```

```{r}

```
