---
title: "3_chi_sq"
author: "Maiko Hata"
format: pdf
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
```

### chi 1: race_us

--- omnibus chi

```{r}
race_us <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/race_us.csv")

race_oregon <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/race_oregon.csv")

```

```{r}
chi_omni <-chisq.test(race_us[, 3:8])
omni_p_value <- chi_omni$p.value
omni_statistic <- chi_omni$statistic 
```

```{r}
chisq.test(race_oregon[, 3:8])
```

### chi 2: byrace

--- US and Oregon entire table. Probably no need as I'll be running the chi-square with residuals anyway. —

```{r}
byrace <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/byrace.csv")

data_oregon <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/data_oregon.csv")
```

```{r}
chisq.test(data_oregon[, 2])
```

### chi 2: agg_by_area

--- withdrawn category —

```{r}
agg_by_area <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/agg_by_area.csv")
```

```{r}
chisq.test(agg_by_area[, 2:3])
```

### chi 3: us_data_attempts_BLWH

--- DQ for BLACK and WHITE for chi-square and odds ratio

```{r}
us_data_attempts <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/us_data_attempts_BLWH.csv")

us_data_attempts_BLWH <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/us_data_attempts_BLWH.csv")

```

```{r}
 
chi_dq_BLWH_result <- chisq.test(us_data_attempts_BLWH[, 2:3])
chi_dq_BLWH_result
```

### chi 4: race_matrix

--- I shouldn't need to "call" race_matrix.csv because it's a df but that means I have to recreate it ---


```{r}
# DROPPED AREA but kept the RACE 
race_matrix <- as.matrix(race_us[, -2, drop = FALSE])  
```

```{r}
race_matrix <- as.matrix(race_matrix)
```

# GAH WTH 

```{r}
# Check for NAs
anyNA(race_matrix)  

# Check for negative values
any(race_matrix < 0)  

# Check for infinite values
any(!is.finite(race_matrix))  

```

```{r}
# Error in stats::chisq.test(x, y, ...) :  -- all entries of 'x' must be nonnegative and finite BUT THERE IS NO NEGATIVE OR NA 
chisq <- chisq.test(race_matrix)
```

# this doesn't run, it says Error in corrplot(chisq$residuals, is.cor = FALSE) : Need a matrix or data frame!

```{r}
# corrplot(chisq$residuals, 
#         is.cor = FALSE) 
```

```{r}
chi_result <- chisq.test(race_matrix)
print(chi_result)
```

# Do I need to save somehow chi_results$stdres as a df...? 

```{r}
print(chi_result$stdres)
# Residual > +2 → Observed count is higher than expected.
# Residual < -2 → Observ
```

# There used to be a Var1 and Var2 now there's only Var1 
```{r}
colnames(residual_matrix)
rownames(residual_matrix)
```

it's NULL so I need to assign names first it says 

```{r}
```

```{r}
residual_matrix <- as.matrix(chisq$residuals)
col_order <- order(colnames(residual_matrix))
row_order <- order(rownames(residual_matrix))
residual_matrix <- residual_matrix[row_order, col_order]
```


```{r}
residuals <- chi_result$stdres

```


```{r}
print(residuals)
```


```{r}
corrplot(residuals,
         is.corr = FALSE)
```

```{ed count is lower than expected.}
```

### chi 5: residuals_df

```{r}
# I'm trying out to see if I can do the chi-square with residuals (per https://chatgpt.com/share/67a1833d-9fc4-8012-8193-b6fc358a9687)
```

```{r}
residuals 
```

```{r}
residuals_df <- as.data.frame(as.table(chi_result$stdres))
```

```{r}
write.csv(residuals_df, 
          file = "/Users/hata/Desktop/EDLD652_Diss/Data/residuals_df.csv", 
          row.names = FALSE)
```

```{r}
# Maybe I shouldn't rename them here? 
# rownames(race_matrix) <- c(
#  "Alaska Native/American Indian", 
#  "Asian", 
#  "Black/African American", 
#  "Hispanic/Latino", 
#  "More than Two Races", 
#  "Pacific Islander", 
#  "White" ) 
```

```{r}
colnames(race_matrix)
```

```{r}
# dropping the "area" column 
race_matrix <- race_matrix[, -1]
```

# DOES NOT WORK WHY

```{r}
# ncol(race_matrix)
```

```{r}
race_matrix <- race_matrix[, 1:6]
colnames(race_matrix) <- c(
  "Not Determined", 
  "Disqualified", 
  "Moved Out", 
  "Part B Eligible", 
  "Not Eligible", 
  "Withdrawn")
```

## Idoesn't render but it almost does - it's rendering pdf and stops. What's wrong with the following chunk?

```{r}
# race_matrix <- race_matrix[, order(colnames(race_matrix))]
```

```{r}
chi_result <- chisq.test(race_matrix)
```

```{r}
print(chi_result$stdres)
```

```{r}

```
