---
title: "4_plots"
author: "Maiko Hata"
format: pdf
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
```

### plot 1: us_data_wide

--- DQ data in US —-

```{r}
us_data_wide <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/us_data_wide.csv")
```

# category names changed already to cleaner names - check the DF so it runs

```{r}
# figure 1 for DQ data in US   

us_data_wide <- us_data_wide %>% 
  select(Race, attempts_to_contact_unsuccessful) %>% 
    mutate(attempts_to_contact_unsuccessful = attempts_to_contact_unsuccessful * 100)
```

```{r}
# figure 1 continues 

ggplot(us_data_wide, aes(x = Race, y = attempts_to_contact_unsuccessful)) +
  geom_col(stat = "identity", fill = "gray43", position = 'dodge') +
   geom_text(aes(label = attempts_to_contact_unsuccessful),
             position = position_dodge(0.9),
             hjust = -0.05,
             size = 2)+
  labs(
    title = "Figure 1: Unsuccessful Contacts/DQ (U.S.)", 
    x = "Race", 
    y = "Percentage of Exit by Race (%)"
  ) +
  theme_classic()+
  scale_x_discrete(labels = c("Alaska Native/American Indian", "Asian", "Black/African American", "Hispanic", "Multiracial", "Pacific Islander", "White"))+
  coord_flip()
```

```{r}
# need to reorder alphabetically the categories 
```

### plot 2: oregon_data_wide_DQ

```{r}
oregon_data_wide_DQ <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/oregon_data_wide_DQ.csv")
```

```{r}
 ggplot(oregon_data_wide_DQ, 
        aes(x = race, y = attempts_to_contact_unsuccessful)) +
  geom_col(stat = "identity", fill = "gray24", position = 'dodge') +
   geom_text(aes(label = attempts_to_contact_unsuccessful),
             position = position_dodge(0.9),
             hjust = -0.3,
             size = 2.2) +
  labs(
    title = "Figure 2: Unsuccessful Contacts/DQ (Oregon)", 
    x = "Race", 
    y = "Percentage of Exit by Race"
  ) +
  theme_classic() +
  scale_x_discrete(labels = c("Alaska Native/American Indian", "Asian", "Black/African American", "Hispanic", "Multiracial", "Pacific Islander", "White"))+
  coord_flip()
 
 # print(oregon_data_wide_DQ) 
```

```{r}
# need to reorder alphabetically the categories 
```

### kable/plot 3: cc_transposed.csv

--- Kable table of CHILDCOUNT in US/OR

```{r}
race_us_chart <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/cc_transposed.csv")
```

```{r}
kable(cc_transposed, 
      col.names = NULL, 
      caption = "Child Count (US \\& Oregon)") %>% 
  column_spec(1, width = "8cm") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, 
                position = "left", 
                font_size = 10.5, 
                latex_options = "scale_down")
```

### plot 4: childcount1920USOR_long

--- ggplots for US & OR Races in CHILDCOUNT

```{r}
race_us_chart <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/childcount1920USOR_long.csv")
```

```{r}
# CAMERON helped - change the race category names back to what it should look like. Don't label them each.

ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) + 
  # HERE"S THE CHANGES
  scale_fill_discrete(labels = c("American Indian")) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  theme_minimal() 
```

```{r}
# Trying to specify group_colors like Cameron taught me. I found this page where they talk about rcartocolor which has color combos https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible The Below codes DO NOT RUN

# 2/14/25DOES NOT RUN BECAUSE THE NAMES IN THE childcount1920USOR_long DOESN'T FIT THE NAMES OF THOSE RACES (below are the full spell, the DF still has clean names)
```

```{r}
# 2/14/25DOES NOT RUN BECAUSE THE NAMES IN THE childcount1920USOR_long DOESN'T FIT THE NAMES OF THOSE RACES (below are the full spell, the DF still has clean names)
# group_colors <- c(
#  "American Indian or Alaska Native" = "#E69F00",
#  "Asian" = "#56B4E9",
#  "Black or African American" = "#009E73",
#  "Hispanic or Latino" = "#F0E442",
#  "Native Hawaiian or Pacific Islander" = "#0072B2",
#  "Two or More Races" = "#D55E00",
#  "White" = "#CC79A7")
```

```{r}
# ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
#  geom_bar(stat = "identity") + 
#  scale_fill_manual(values = group_colors) +  
#  scale_y_continuous(labels = percent_format(scale = 1)) +
#  labs(title = "Percentage of Child Count by Race in US & Oregon",
#       x = "State",
#       y = "Percentage",
#       fill = "Race") +
#  theme_minimal()
```

```{r}
# Example of HORRIBLE work 
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = race), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 4) +
  theme_minimal()
```

## plot 5: residuals_df

--- Chi-square with Residuals —

```{r}
residuals_df <- read.csv("/Users/hata/Desktop/EDLD652_Diss/Data/residuals_df.csv
```

## Caused by error: ! object 'Var2' not found Run `rlang::last_trace()` to see where the error occurred. \>

```{r}
# because of Caused by error:! object 'Var2' not found Run `rlang::last_trace()` to see where the error occurred., I deleted Var2, after Var1, 

ggplot(residuals_df, aes(Var1, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", 
    mid = "white", 
    high = "red") +
  labs(
    title = "Chi-Square Standardized Residuals",
    x = "Race", 
    y = "Exit Reasons") +
  theme_minimal()
```

```{r}

```
