---
title: "4_plots"
author: "Maiko Hata"
format: pdf
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
```

### kable/plot 3: cc_transposed.csv

--- Kable table of CHILDCOUNT in US/OR

```{r}
cc_transposed <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/cc_transposed.csv")
```

```{r}
kable(cc_transposed, 
      col.names = NULL, 
      caption = "Child Count (US \\& Oregon)") %>% 
  column_spec(1, width = "8cm") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, 
                position = "left", 
                font_size = 10.5, 
                latex_options = "scale_down")
```

### plot 4: childcount1920USOR_long

--- ggplots for US & OR Races in CHILDCOUNT

```{r}
childcount1920USOR_long <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/childcount1920USOR_long.csv")
```

```{r}
print(childcount1920USOR_long)
```

### CAMERON, I need to change the row names to race names

```{r}
childcount1920USOR_long <- childcount1920USOR_long %>%
  mutate(race = recode(race,
         "white_percent" = "White",
         "black_or_african_american_percent" = "Black/African American",
         "asian_percent" = "Asian",
         "hispanic_latino_percent" = "Hispanic/Latino",
         "american_indian_or_alaska_native_percent" = "American Indian/Alaska Native",
         "native_hawaiian_or_pacific_islander_percent" = "Native Hawaiian/Other Pacific Islander",
         "two_or_more_races_percent" = "Two or More Races"))

```

Rewrite CSV with the race names corrected!!

```{r}
write.csv(childcount1920USOR_long, file = "/Users/hata/Desktop/EDLD652_Diss/Data/childcount1920USOR_long.csv",           row.names = FALSE)

```

```{r}
childcount1920USOR_long <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/childcount1920USOR_long.csv")

```

## CAMERON, Why does this one NOT show the race titles somehow?

```{r}
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) + 
  # HERE"S THE CHANGES
  scale_fill_discrete(labels = c("American Indian")) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  theme_minimal() 
```

```{r}
# Trying to specify group_colors like Cameron taught me. I found this page where they talk about rcartocolor which has color combos https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible The Below codes DO NOT RUN

# 2/14/25DOES NOT RUN BECAUSE THE NAMES IN THE childcount1920USOR_long DOESN'T FIT THE NAMES OF THOSE RACES (below are the full spell, the DF still has clean names)
```

```{r}
# 2/14/25DOES NOT RUN BECAUSE THE NAMES IN THE childcount1920USOR_long DOESN'T FIT THE NAMES OF THOSE RACES (below are the full spell, the DF still has clean names)
# group_colors <- c(
#  "American Indian or Alaska Native" = "#E69F00",
#  "Asian" = "#56B4E9",
#  "Black or African American" = "#009E73",
#  "Hispanic or Latino" = "#F0E442",
#  "Native Hawaiian or Pacific Islander" = "#0072B2",
#  "Two or More Races" = "#D55E00",
#  "White" = "#CC79A7")
```

```{r}
# ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
#  geom_bar(stat = "identity") + 
#  scale_fill_manual(values = group_colors) +  
#  scale_y_continuous(labels = percent_format(scale = 1)) +
#  labs(title = "Percentage of Child Count by Race in US & Oregon",
#       x = "State",
#       y = "Percentage",
#       fill = "Race") +
#  theme_minimal()
```

```{r}
# Example of HORRIBLE work 
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = race), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 4) +
  theme_minimal()
```

```{r}
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +  # "dodge" makes it a grouped bar chart
  scale_fill_viridis_d() +  # Apply Viridis discrete color palette
  scale_y_continuous(labels = percent_format(scale = 1)) +  # Format y-axis as percentages
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  theme_minimal() 
```

```{r}
ggplot(childcount1920USOR_long, 
       aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +  
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1)) +  
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = race), 
            color = "black", 
            size = 4) +
  theme_minimal() 
```

```{r}
# not good - OR vs US side by side 
ggplot(childcount1920USOR_long, 
       aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.05))) +  
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5,  # Moves text slightly above bars
            color = "black", 
            size = 4) +
  theme_minimal(base_size = 14) +  
  theme(legend.position = "bottom")  
```

```{r}
ggplot(childcount1920USOR_long, aes(x = race, y = percent, fill = state)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  # Side-by-side bars
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "Race",
       y = "Percentage",
       fill = "State") +
  geom_text(aes(label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5,  
            color = "black", 
            size = 4) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")  # Moves legend below for clarity
```

```{r}
# cameron's favorite
ggplot(childcount1920USOR_long, 
       aes(x = percent, y = race, fill = state)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  # Side-by-side bars
  scale_fill_viridis_d() +  
  scale_x_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "Percentage",
       y = "Race",
       fill = "State") +
  geom_text(aes(label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            hjust = -0.2,  
            color = "black", 
            size = 4) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")  # Moves legend below for clarity

```

trying to show in times new roman

```{r}
library(extrafont)
font_import(pattern = "Times", 
            prompt = FALSE)
loadfonts(device = "win")
```

```{r}
ggplot(childcount1920USOR_long, 
       aes(x = percent, 
           y = race, 
           fill = state)) +
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.7)) +  
  scale_fill_viridis_d() +  
  scale_x_continuous(
    labels = percent_format(scale = 1), 
    expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "Percentage",
       y = "Race",
       fill = "State") +
  geom_text(aes(
    label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            hjust = -0.2,  
            color = "black", 
            size = 4, 
            family = "Times New Roman") +  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times New Roman")  
  )
```

trying to change the colors to okabe ito

```{r}
library(ggsci)
```

```{r}
ggplot(childcount1920USOR_long, 
       aes(x = percent, 
           y = race, 
           fill = state)) +
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.7)) +  # ✅ Closed geom_bar() correctly
  scale_fill_aaas() +  # ✅ Correct usage of Okabe-Ito colors
  scale_x_continuous(
    labels = percent_format(scale = 1)  # ✅ Removed misplaced labs() call
  ) +  
  labs(
    title = "Percentage of Child Count by Race in US & Oregon",
    x = "Percentage",
    y = "Race",
    fill = "State"
  ) +  
  geom_text(aes(
    label = scales::percent(percent / 100, accuracy = 0.1)), 
    position = position_dodge(width = 0.7), 
    hjust = -0.2,  
    color = "black", 
    size = 4, 
    family = "Times New Roman"  
  ) +  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times New Roman")  # ✅ Correctly closed theme()
  )
```

trying to change the color of the labels but it LOSES the state label for the US below somehow. WHY!?!?

```{r}
palette.colors(palette = "Okabe-Ito")
##         black        orange       skyblue   bluishgreen        yellow 
##     "#000000"     "#E69F00"     "#56B4E9"     "#009E73"     "#F0E442" 
##          blue    vermillion reddishpurple          gray 
##     "#0072B2"     "#D55E00"     "#CC79A7"     "#999999" 
```

```{r}
# Load the ggsci package for Okabe-Ito palette
library(ggsci) 
```

```{r}
# Assign colors from Okabe-Ito palette by name
# NO TIMES NEW ROMAN for EDLD652 final project 
okabe_ito_colors <- c("US" = pal_aaas("default")(10)[6],      
                      # Yellow (#F0E442)
                      "Oregon" = pal_aaas("default")(10)[3])  
                      # Bluish green (#009E73)

# Plot with Okabe-Ito colors
ggplot(childcount1920USOR_long, 
       aes(x = percent, 
           y = race, 
           fill = state)) +
  geom_bar(stat = "identity", 
           position = position_dodge(
          width = 0.7)) +  
  scale_fill_manual(
    values = okabe_ito_colors, 
    breaks = c("US", "Oregon"),         
    labels = c("US", "Oregon")          
  ) +
  scale_x_continuous(
    labels = percent_format(scale = 1), 
    expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Percentage of Child Count by Race in US & Oregon",
    x = "Percentage",
    y = "Race",
    fill = "State"
  ) +
  geom_text(aes(
    label = scales::percent(percent / 100, accuracy = 0.1)), 
    position = position_dodge(width = 0.7), 
    hjust = -0.2,  
    color = "black", 
    size = 4, 
    family = "Times New Roman"  
  ) +  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom"
  )
```

```{r}
# Assign colors from Okabe-Ito palette by name
okabe_ito_colors <- c("US" = pal_aaas("default")(10)[6],      
                      # Yellow (#F0E442)
                      "Oregon" = pal_aaas("default")(10)[3])  
                      # Bluish green (#009E73)

# Plot with Okabe-Ito colors
ggplot(childcount1920USOR_long, 
       aes(x = percent, 
           y = race, 
           fill = state)) +
  geom_bar(stat = "identity", 
           position = position_dodge(
          width = 0.7)) +  
  scale_fill_manual(
    values = okabe_ito_colors, 
    breaks = c("US", "Oregon"),         
    labels = c("US", "Oregon")          
  ) +
  scale_x_continuous(
    labels = percent_format(scale = 1), 
    expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Percentage of Child Count by Race in US & Oregon",
    x = "Percentage",
    y = "Race",
    fill = "State"
  ) +
  geom_text(aes(
    label = scales::percent(percent / 100, accuracy = 0.1)), 
    position = position_dodge(width = 0.7), 
    hjust = -0.2,  
    color = "black", 
    size = 4, 
    family = "Times New Roman"  
  ) +  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Times New Roman")  
  )
```

```{r}

```
