---
title: "4_plots_or_USRACE_average"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
library(epitools)
library(vcd)
library(knitr)
```

```{r}
us_data_DQ_proportion_w_average <- 
  read.csv("/Users/hata/Desktop/EDLD652_Diss/Data/us_data_DQ_proportion_w_average.csv")
```

```{r}
us_data_DQ_proportion_w_average
```

3/29/25: This following chunks don't run (ChatGPT fail), but could be a great one for functional programming class ideas?

```{r}
# ChatGPT: 
# 1) library(vcd) 
# 2) Define function to compare row i with the Average row (row 8) 
# compare_to_average <- function(i) {
# 3) extract the two rows (as a 2x2 matrix)
# temp_table <- rbind(us_data_DQ_proportion_w_average[i, 2:3], 
#                    us_data_DQ_proportion_w_average[8, 2:3])
# compute odds ratio 
# or <- oddsratio(as.matrix(temp_table), log = FALSE)

#return as data frame 
# data.frame(race = us_data_DQ_proportion_w_average[i, "race"], 
#           OddsRatio = or$measure[1])}
```

```{r}
# us_oddsratio_results <- do.call(rbind, lapply(1:7, compare_to_average))
# print(results)
```

Modifying df us_data_DQ_w_average

```{r}
#The $ operator in R is used to access a specific column of a data frame (or a named list). It's one of the most common and useful operators in R.
us_data_DQ_proportion_w_average$OtherExits <- 
  us_data_DQ_proportion_w_average$exit_total - 
  us_data_DQ_proportion_w_average$attempts_to_contact_unsuccessful
```

Trying to run each of the Odds Ratio combo by hand using us_data_DQ_proportion_w_average df, but learned that for Odds Ratio, I have to use contingency table of Dismissal vs OTHER EXITS. Oops. See above for a new df of dismissal but with a new column.

```{r}
library(vcd)

# Row 1 = Group 1, Row 8 = Average
# a <- us_data_DQ_proportion_w_average[1, "attempts_to_contact_unsuccessful"]
# b <- us_data_DQ_proportion_w_average[1, "OtherExits"]
# c <- us_data_DQ_proportion_w_average[8, "attempts_to_contact_unsuccessful"]
# d <- us_data_DQ_proportion_w_average[8, "OtherExits"]

# 2x2 table
# temp_table <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE)

# Check table
# print(temp_table)

# Calculate odds ratio
# oddsratio(temp_table, log = FALSE)
```

```{r}
us_race_3cols_8rows <- us_data_DQ_proportion_w_average %>% 
  rename(Dismissed = attempts_to_contact_unsuccessful)

```

```{r}
us_race_3cols_8rows <- us_race_3cols_8rows %>% 
  select(2, 4, 6)
```

```{r}
write.csv(us_race_3cols_8rows, 
          file="/Users/hata/Desktop/EDLD652_Diss/Data/us_race_3cols_8rows.csv")
```

Calculating one Odds Ratio by one but using the us_race_3cols_8rows to minimize my pain time

```{r}
library(vcd)

# Example: Group 1 vs Average (row 1 vs row 8)
# temp_table <- as.matrix(rbind(us_race_3cols_8rows[1, c("Dismissed", "OtherExits")],
#                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))

#oddsratio(temp_table, log = TRUE)
# THIS ONLY RETURNS THE ODDS RATIO SO I"M SWTICHING TO EPITOOLS WITH WALD METHOD
```

```{r}
library(epitools)

# Example: Group 1 vs Average (row 1 vs row 8)
temp_table <- as.matrix(rbind(us_race_3cols_8rows[1, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))

# "wald" is the default method for odds ratio with confidence intervals. 
epitools::oddsratio(temp_table, method = "wald")
```

Results from above - Row 1 AM vs Average:

odds ratio with 95% C.I. estimate lower upper 1.876783 1.806008 1.950332

```{r}
temp_table <- as.matrix(rbind(us_race_3cols_8rows[2, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 2: AS vs Average: odds ratio with 95% C.I. estimate lower upper 0.5543598 0.5396222 0.5694998

```{r}
temp_table <- as.matrix(rbind(us_race_3cols_8rows[3, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))
epitools::oddsratio(temp_table, method = "wald")
```

Row 3: BL vs Average: odds ratio with 95% C.I. estimate lower upper 1.843126 1.817768 1.868837

```{r}
temp_table <- as.matrix(rbind(us_race_3cols_8rows[4, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))
epitools::oddsratio(temp_table, method = "wald")
```

Row 4: HI vs Average: odds ratio with 95% C.I. estimate lower upper 1.142779 1.12817 1.157576

```{r}
temp_table <- as.matrix(rbind(us_race_3cols_8rows[5, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))
epitools::oddsratio(temp_table, method = "wald")
```

Row 5: MU vs Average: odds ratio with 95% C.I. estimate lower upper 1.110257 1.086377 1.134662

```{r}
temp_table <- as.matrix(rbind(us_race_3cols_8rows[6, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))
epitools::oddsratio(temp_table, method = "wald")
```

Row 6: PI vs Average: odds ratio with 95% C.I. estimate lower upper 1.192017 1.115164 1.274168

```{r}
temp_table <- as.matrix(rbind(us_race_3cols_8rows[7, c("Dismissed", "OtherExits")],
                              us_race_3cols_8rows[8, c("Dismissed", "OtherExits")]))
epitools::oddsratio(temp_table, method = "wald")
```

Row 7: WH vs Average: odds ratio with 95% C.I. estimate lower upper 0.7503513 0.7411056 0.7597124

```{r}
us_race_odds_ratio_vs_average <- data.frame(
  Race = c("American Indian/Alaska Native", 
           "Asian", 
           "Black/African American", 
           "Hispanic", 
           "Native Hawaiian and Other Pacific Islander", 
           "Two or More Races", 
           "White"), 
  Odds_Ratio = c(
    1.876783, 
    0.5543598, 
    1.843126, 
    1.142779, 
    1.192017, 
    1.110257, 
    0.7503513
  ), 
  ci_lower = c(
    1.806008, 
0.5396222, 
1.817768, 
1.12817, 
1.115164, 
1.086377, 
0.7411056
     ), 
  ci_upper = c(
    1.950332, 
0.5694998, 
1.868837, 
1.157576, 
1.274168, 
1.134662, 
0.7597124
    )
)
```

```{r}
write.csv(us_race_odds_ratio_vs_average, 
          file="/Users/hata/Desktop/EDLD652_Diss/Data/us_race_odds_ratio_vs_average.csv")
```

```{r}
kable(us_race_odds_ratio_vs_average, 
      caption = "Odds Ratio: National Average")
```

```{r}
# Flipped the order on X axis AND added the line for "1" 
ggplot(us_race_odds_ratio_vs_average, 
       aes(x = reorder(Race, Odds_Ratio), y = Odds_Ratio)) + 
  geom_col(fill = "lightgray") +  # this is now fixed
  coord_flip() +
  geom_hline(
    yintercept = 1, 
    linetype = "dashed", 
    color = "green") +  
  scale_x_discrete(limits = rev(sort(unique(us_race_odds_ratio_vs_average$Race)))) + 
  labs(
    title = "Odds Ratio for Dismissal (Compared to Average)", 
    x = "Race", 
    y = "Odds Ratio"
  ) + 
  theme_minimal()

```

### Final Forrest Plot for US RACE against national average!!! Congratulations, Maiko 🎉㊗🥳

```{r}
ggplot(us_race_odds_ratio_vs_average, 
       aes(x = reorder(Race, Odds_Ratio), y = Odds_Ratio)) + 
  geom_point(size = 4, color = "skyblue") +  # OR point
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # CI bars
  geom_text(aes(label = round(Odds_Ratio, 2)), 
            vjust = -0.8, size = 3.5, family = "Times New Roman") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "green") +  # Reference line
  coord_flip() +  # Flip axes
  scale_x_discrete(limits = rev(sort(unique(us_race_odds_ratio_vs_average$Race)))) + 
  labs(
    title = "Odds Ratio for Dismissal", 
    x = "Group", 
    y = "Odds Ratio"
  ) + 
  theme_minimal() +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold")
  )

```
