---
title: "4_plots_or"
author: "Maiko Hata"
format: 
  pdf:
    mainfont: "Times New Roman"
    sansfont: "Times New Roman"
fig-width: 7
fig-height: 4
csl: apa.csl
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
library(epitools)
```

### odd ratio 1: us_data_attempts_BLWH

--- OR for black vs white DQ —

```{r}
us_data_attempts_BLWH <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/us_data_attempts_BLWH.csv")
```

```{r}
# chi_dq_BLWH_result <- chisq.test(us_data_attempts_BLWH[, 2:3])
```

```{r}
# print(chi_dq_BLWH_result)
# data:  us_data_attempts_BLWH[, 2:3]
# X-squared = 22556, df = 1, p-value < 2.2e-16
```

```{r}
# used wording from Morgan et al. (2020) and ChatGPT 
# prepping the df to mutate a column to have non_dq and delete total 

us_data_wide_DQ_BLWH <- us_data_attempts_BLWH %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)

us_data_wide_DQ_BLWH_nototal <- us_data_wide_DQ_BLWH %>% 
  select(-exit_total)
```

```{r}
#Per chatGPT. The columns and rows switched...? 
us_dq_BLWH_matrix <- matrix( 
  c(56155, 359718, 99922, 1572263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("Black", "White")
  ))
```

```{r}
# ODDS RATIO SUCCESS 
oddsratio(us_dq_BLWH_matrix)
```

```{r}
dq_black_OR <- oddsratio(us_dq_BLWH_matrix)
```

### odd ratio 2:

```{r}
byrace <- read_csv("/Users/hata/Desktop/EDLD652_Diss/Data/byrace.csv")
```

```{r}
# remaking a df with exit total and DQ 12/1
us_data_attempts <- byrace %>% 
  filter(area == "US and Outlying Areas") %>% 
  group_by(race) %>% 
  summarize(exit_total = sum(exiting_total2), 
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful)) %>% 
  mutate(proportion = attempts_to_contact_unsuccessful / exit_total)
```

```{r}
write.csv(us_data_attempts, file = "/Users/hata/Desktop/EDLD652_Diss/Data/us_data_attempts.csv",row.names = FALSE)
```

### odds ratio 3: dq_native

--- Odds Ratio for Native American vs White DQ —-\
see EDLD652 Data Visualization notes google doc for interpretations

```{r}
dq_native_us <- us_data_attempts %>% 
  select(race, exit_total, attempts_to_contact_unsuccessful) %>% 
filter(race == "AM" | race == "WH") 
```

```{r}
dq_native_us_nototal <- dq_native_us %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)
```

```{r}
dq_native_us_nototal <- dq_native_us_nototal %>% 
  select(-exit_total)
```

```{r}
dq_native_us_matrix <- matrix( 
  c(3262, 20521, 99922, 1572263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("Alaska Native", "White")
  ))
```

```{r}
  oddsratio(dq_native_us_matrix)
```

```{r}
dq_native_us_OR <- oddsratio(dq_native_us_matrix)
```

```{r}
write.csv(dq_native_us, file = "/Users/hata/Desktop/EDLD652_Diss/Data/dq_native_us.csv", row.names = FALSE)
```

```{r}
chi_dq_native <- chisq.test(dq_native_us[, 2:3])
```

```{r}
print(chi_dq_native)
# data:  us_data_attempts_BLWH[, 2:3]
# X-squared = 22556, df = 1, p-value < 2.2e-16
```

### odds ratio 4: dq_hispanic

--- Odds Ratio for Hispanic vs White DQ —-\
see EDLD652 Data Visualization notes google doc for interpretations

```{r}
dq_hispanic_us <- us_data_attempts %>% 
  select(race, exit_total, attempts_to_contact_unsuccessful) %>% 
filter(race == "HI" | race == "WH") 
```

```{r}
write.csv(dq_hispanic_us, file = "/Users/hata/Desktop/EDLD652_Diss/Data/dq_hispanic_us.csv", 
          row.names = FALSE)
```

```{r}
dq_hispanic_us_nototal <- dq_hispanic_us %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)
```

```{r}
dq_hispanic_us_nototal <- dq_hispanic_us_nototal %>% 
  select(-exit_total)
```

```{r}
dq_hispanic_us_matrix <- matrix( 
  c(79951, 826020, 99922, 1672263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("Hispanic/Latino", "White")
  ))
```

```{r}
  oddsratio(dq_hispanic_us_matrix)
```

```{r}
 dq_hispanic_us_OR <- oddsratio(dq_hispanic_us_matrix)
```

### odds ratio 5: dq_asian

--- Odds Ratio for Asian vs White DQ —-\
see EDLD652 Data Visualization notes google doc for interpretations

```{r}
dq_asian_us <- us_data_attempts %>% 
  select(race, exit_total, attempts_to_contact_unsuccessful) %>% 
filter(race == "AS" | race == "WH") 
```

```{r}
dq_asian_us_nototal <- dq_asian_us %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)
```

```{r}
dq_asian_us_nototal <- dq_asian_us_nototal %>% 
  select(-exit_total)
```

```{r}
dq_asian_us_matrix <- matrix( 
  c(6559, 139693, 99922, 1572263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("Asian", "White")
  ))
```

```{r}
oddsratio(dq_asian_us_matrix)
```

```{r}
dq_asian_us_OR <- oddsratio(dq_asian_us_matrix)
```

### Odds ratio 6: dq_multi

--- Odds Ratio for \_\_\_ vs White DQ —-\
see EDLD652 Data Visualization notes google doc for interpretations

```{r}
dq_multi_us <- us_data_attempts %>% 
  select(race, exit_total, attempts_to_contact_unsuccessful) %>% 
filter(race == "MU" | race == "WH") 
```

```{r}
dq_multi_us_nototal <- dq_multi_us %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)
```

```{r}
dq_multi_us_nototal <- dq_multi_us_nototal %>% 
  select(-exit_total)
```

```{r}
dq_multi_us_matrix <- matrix( 
  c(11676, 124165, 99922, 1572263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("More than Two Races", "White")
  ))
```

```{r}
oddsratio(dq_multi_us_matrix)
```

```{r}
dq_multi_us_OR <- oddsratio(dq_multi_us_matrix)
```

### odds ratio 5: dq_pi

--- Odds Ratio for Pacific Islaner vs White DQ —-\
see EDLD652 Data Visualization notes google doc for interpretations

```{r}
dq_pi_us <- us_data_attempts %>% 
  select(race, exit_total, attempts_to_contact_unsuccessful) %>% 
filter(race == "PI" | race == "WH") 
```

```{r}
dq_pi_us_nototal <- dq_pi_us %>% 
  mutate(non_dq = exit_total - attempts_to_contact_unsuccessful)
```

```{r}
dq_pi_us_nototal <- dq_pi_us_nototal %>% 
  select(-exit_total)
```

```{r}
dq_pi_us_matrix <- matrix(
  c(977, 9677, 99922, 1572263), 
  nrow = 2, 
  dimnames = list(
    c("attempts_to_contact_unsuccessful", "Other"), 
    c("Alaska Native", "White")
  ))
```

```{r}
oddsratio(dq_pi_us_matrix)
```

```{r}
dq_pi_us_OR <- oddsratio(dq_pi_us_matrix)
```

### Odds ratio plot 1:

```{r}
or_us_white <- data.frame(
  group = c(
    "American Indian/Alaska Native", 
    "Asian", 
    "Black/African American", 
    "Hispanic/Latino", 
    "More than Two Races", 
    "Pacific Islander", 
    "White (Reference)"),
  odds_ratio = c(
    2.501309, 
    0.7388185, 
    2.456352, 
    1.619882, 
    1.479668, 
    1.588854, 
    1)
)
```

v1

```{r}
ggplot(or_us_white, 
       aes(x = factor(group, 
                      levels = sort(unique(group))), 
           y = odds_ratio)) +
  geom_col(fill = "skyblue") +
  geom_hline(
    yintercept = 1, 
    linetype = "dashed", 
    color = "red") +  
  labs(title = "Odds Ratios Compared to White (Reference)", 
       x = "Race", 
       y = "Odds Ratio") +
  theme_minimal()
```

v2: flipped coord

```{r}
# with flipped coordinates
  ggplot(or_us_white, 
         aes(
           x = factor(
             group, 
             levels = sort(unique(group))), 
           y = odds_ratio)) +
  geom_col(
    fill = "skyblue") +
  geom_hline(
    yintercept = 1, 
    linetype = "dashed", 
    color = "red") +  
  coord_flip() +  
  labs(
    title = "Odds Ratios Compared to White (Reference)", 
    x = "Group", 
    y = "Odds Ratio"
  ) +
  theme_minimal()
```

```{r}
# I need to alphabetize 'group' along the y-axis

```

```{r}
or_us_white_ci <- data.frame(
  group = c(
    "American Indian/Alaska Native", 
    "Asian", 
    "Black/African American", 
    "Hispanic/Latino", 
    "More than Two Races", 
    "Pacific Islander", 
    "White (Reference)"),
  odds_ratio = c(
    2.501309, 
    0.7388185, 
    2.456352, 
    1.619882, 
    1.479668, 
    1.588854, 
    1), 
  ci_lower = c(
    2.408865, 
    0.7200855, 
    2.429572, 
   1.60425, 
   1.450269, 
   1.486348, 
   0.950000), 
  ci_upper = c(
   2.596393, 
  0.7578737, 
  2.483474, 
  1.635544, 
  1.50953, 
  1.696447, 
  1.050000)
)
```

Forrest Plot 1

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, 
           y = odds_ratio)) +
  geom_point(size = 3, 
             color = "skyblue") +  
  geom_errorbar(aes(
    ymin = ci_lower, 
    ymax = ci_upper), 
    width = 0.2) +  
  geom_hline(
    yintercept = 1, 
    linetype = "dashed", 
    color = "red") +  
  coord_flip() +  
  labs(
    title = "Forest Plot of Odds Ratios",
    x = "Race",
    y = "Odds Ratio"
  ) +
  theme_minimal()
```

trying to reverse alphaetize the Y axis

```{r}
or_us_white_ci$group <- factor(or_us_white_ci$group, levels = rev(sort(unique(or_us_white_ci$group))))

```

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, y = odds_ratio)) +
  geom_point(size = 3, color = "skyblue") +  
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +  
  coord_flip() +  
  labs(
    title = "Forest Plot of Odds Ratios",
    x = "Race",
    y = "Odds Ratio"
  ) +
  theme_minimal()
```

made the dots bigger

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, y = odds_ratio)) +
  geom_point(size = 5, color = "skyblue") +  # Increased size from 3 to 5
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +  
  coord_flip() +  
  labs(
    title = "Forest Plot of Odds Ratios",
    x = "Race",
    y = "Odds Ratio"
  ) +
  theme_minimal()
```

added the data label yeaaaah

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, 
           y = odds_ratio)) +
  geom_point(size = 4, 
             color = "skyblue") +  
  geom_errorbar(aes(
    ymin = ci_lower, 
    ymax = ci_upper), 
    width = 0.2) +  
  geom_text(aes(
    label = round(odds_ratio, 2)), 
    vjust = -0.2, size = 3.5) +  
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "red") +  
  coord_flip() +  
  labs(
    title = "Forest Plot of Odds Ratios",
    x = "Group",
    y = "Odds Ratio"
  ) +
  theme_minimal()

```

moved the labels above the dots

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, 
           y = odds_ratio)) +
  geom_point(size = 4, 
             color = "skyblue") +  
  geom_errorbar(aes(
    ymin = ci_lower, 
    ymax = ci_upper), 
    width = 0.2) +  
  geom_text(aes(
    label = round(odds_ratio, 2)), 
    vjust = -0.8, size = 3.5) +  
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "red") +  
  coord_flip() +  
  labs(
    title = "Odds Ratio for Disqulification",
    x = "Group",
    y = "Odds Ratio") +
  theme_minimal() 
```

made the labels for axises bigger

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, 
           y = odds_ratio)) +
  geom_point(size = 4, 
             color = "skyblue") +  
  geom_errorbar(aes(
    ymin = ci_lower, 
    ymax = ci_upper), 
    width = 0.2) +  
  geom_text(aes(
    label = round(odds_ratio, 2)), 
    vjust = -0.8, size = 3.5) +  
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "red") +  
  coord_flip() +  
  labs(
    title = "Odds Ratio for Disqulification",
    x = "Group",
    y = "Odds Ratio") +
  theme_minimal() + 
theme(
    axis.text.x = element_text(size = 10),  
    axis.text.y = element_text(size = 10)   
  )
```

trying to see if I can make it in times new roman

```{r}
ggplot(or_us_white_ci, 
       aes(x = group, 
           y = odds_ratio)) +
  geom_point(size = 4, 
             color = "skyblue") +  
  geom_errorbar(aes(
    ymin = ci_lower, 
    ymax = ci_upper), 
    width = 0.2) +  
  geom_text(aes(
    label = round(odds_ratio, 2)), 
    vjust = -0.8, size = 3.5, 
    family = "Times New Roman") +  
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "red") +  
  coord_flip() +  
  labs(
    title = "Odds Ratio for Disqulification",
    x = "Group",
    y = "Odds Ratio") +
  theme_minimal() + 
theme(
    text = element_text(family = "Times New Roman"),     
    axis.text.x = element_text(size = 10),               
    axis.text.y = element_text(size = 10),               
    plot.title = element_text(size = 12, face = "bold") 
  )
```
