---
title: "4_plots_or_ORRACE_average"
author: "Maiko Hata"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
library(epitools)
library(vcd)
library(knitr)
library(corrplot)
```

```{r}
race_oregon <- read.csv("/Users/hata/Desktop/EDLD652_Diss/Data/race_oregon.csv")
```

```{r}
race_oregon <- race_oregon %>% 
  select(-2, -4, -6, -7, -8, -9)

```

```{r}
race_oregon_w_average <- race_oregon 
```

```{r}
# modifying the df so it has the average row 

race_oregon_w_average <- race_oregon %>% 
  bind_rows(
    race_oregon[1, ] %>%
      mutate(
        across(everything(), ~NA), 
        !!names(.)[1] := "All Race Average",
        !!names(.)[2] := mean(pull(race_oregon, 2), na.rm = TRUE),
        !!names(.)[3] := mean(pull(race_oregon, 3), na.rm = TRUE)
      )
  )

```

```{r}
# Calculate averages first
race_oregon_w_average <- race_oregon %>% 
  bind_rows(
    race_oregon[1, ] %>% 
      mutate(
        across(everything(), ~NA), 
        !!names(.)[1] := "All Race Average", 
        !!names(.)[2] := mean(pull(race_oregon, 2), na.rm = TRUE), 
        !!names(.)[3] := mean(pull(race_oregon, 3), na.rm = TRUE)
      )
  )
```

```{r}
race_oregon_w_average <- race_oregon_w_average %>% 
  rename(Dismissed = attempts_to_contact_unsuccessful)
```

```{r}
# Modigying the df to create OtherExits for odds ratio 
# The $ operator in R is used to access a specific column of a data frame (or a named list). It's one of the most common and useful operators in R.

race_oregon_w_average$OtherExits <- 
  race_oregon_w_average$exit_total -
  race_oregon_w_average$Dismissed
```

```{r}
race_oregon_w_average <- race_oregon_w_average %>% 
  select(-2)
```

```{r}
write.csv(race_oregon_w_average, 
            file="/Users/hata/Desktop/EDLD652_Diss/Data/race_oregon_w_average.csv")
```

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    1, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

# "wald" is the default method for odds ratio with confidence intervals. 
epitools::oddsratio(temp_table, method = "wald")
```

Row 1 AM: odds ratio with 95% C.I. estimate lower upper 1.490971 1.014694 2.190805

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    2, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 2 AS: odds ratio with 95% C.I. estimate lower upper 0.6137758 0.459224 0.8203419

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    3, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 3 BL: odds ratio with 95% C.I. estimate lower upper 1.196284 0.9166426 1.561236

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    4, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 4 HI: odds ratio with 95% C.I. estimate lower upper 1.211377 1.059952 1.384436

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    5, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 5 MU: odds ratio with 95% C.I. estimate lower upper 0.6823019 0.5337505 0.8721976

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    6, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 6 PI: odds ratio with 95% C.I. estimate lower upper 1.270903 0.7084633 2.279858

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    8, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 7 WH: odds ratio with 95% C.I. estimate lower upper 0.9554339 0.8473886 1.077255

```{r}
race_oregon_w_average_full <- data.frame(
  Race = c("American Indian/Alaska Native", 
           "Asian", 
           "Black/African American", 
           "Hispanic", 
           "Native Hawaiian and Other Pacific Islander", 
           "Two or More Races", 
           "White"), 
  Odds_Ratio = c(
    1.490971, 
0.6137758, 
1.196284, 
1.211377, 
1.270903, 
0.6823019, 
0.9554339
), 
  ci_lower = c(
1.014694, 
0.459224, 
0.9166426, 
1.059952, 
0.7084633, 
0.5337505, 
0.8473886
), 
  ci_upper = c(
    2.190805, 
0.8203419, 
1.561236, 
1.384436, 
2.279858, 
0.8721976, 
1.077255
    )
)
```

race_oregon_w_average_full.csv - all the OREGON races' odds ratio against average

```{r}
write.csv(race_oregon_w_average_full, 
          file="/Users/hata/Desktop/EDLD652_Diss/Data/race_oregon_w_average_full.csv")
```

```{r}
kable(race_oregon_w_average_full, 
      caption = "Odds Ratio: Oregon Average")
```

```{r}
ggplot(race_oregon_w_average_full, 
       aes(x = reorder(Race, Odds_Ratio), 
           y = Odds_Ratio)) + 
  geom_point(size = 4, color = "skyblue") + 
  geom_errorbar(aes(ymin = ci_lower, 
                    ymax = ci_upper), 
                width = 0.2) +  
  geom_text(aes(label = round(Odds_Ratio, 2)), 
            vjust = -0.8, 
            size = 3.5, 
            family = "Times New Roman") +
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "green") +  
  coord_flip() +  # Flip axes
  scale_x_discrete(limits = rev(sort(unique(
    race_oregon_w_average_full$Race)))) + 
  labs(
    title = "Odds Ratio for Dismissal (Oregon)", 
    x = "Group", 
    y = "Odds Ratio"
  ) + 
  theme_minimal() +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold")
  )

```

## BELOW are DQ ODDS RATIO against WHITE in Oregon

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    1, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 1 AM: 1.560518 1.07474 2.265865

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    2, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 2 AS: 0.6424053 0.4883897 0.8449903

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    3, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 3 BL: 0.6424053 0.4883897 0.8449903

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    4, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 4 HI: 1.267882 1.154276 1.392668

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    5, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 5 MU: 0.7141278 0.5694253 0.8956023

```{r}
temp_table <- as.matrix(rbind(
  race_oregon_w_average[
    6, c("Dismissed", "OtherExits")],
  race_oregon_w_average[
    7, c("Dismissed", "OtherExits")]))

epitools::oddsratio(temp_table, method = "wald")
```

Row 6 PI: 1.330185 0.7472923 2.367736

```{r}
race_oregon_w_WHITE <- data.frame(
  Race = c("American Indian/Alaska Native", 
           "Asian", 
           "Black/African American", 
           "Hispanic", 
           "Native Hawaiian and Other Pacific Islander", 
           "Two or More Races", 
           "White"), 
  Odds_Ratio = c(
1.560518, 
0.6424053, 
1.252085, 
1.267882, 
1.330185, 
0.7141278, 
1
), 
  ci_lower = c(
1.07474, 
0.4883897, 
0.9763492, 
1.154276, 
0.7472923, 
0.5694253, 
1
), 
  ci_upper = c(
    2.265865, 
0.8449903, 
1.605692, 
1.392668, 
2.367736, 
0.8956023, 
1
    )
)
```

```{r}
write.csv(race_oregon_w_WHITE, 
          file="/Users/hata/Desktop/EDLD652_Diss/Data/race_oregon_w_WHITE.csv")
```

```{r}
kable(race_oregon_w_WHITE, 
      caption = "Odds Ratio Compared to White Group")
```

```{r}
ggplot(race_oregon_w_WHITE, 
       aes(x = reorder(Race, Odds_Ratio), 
           y = Odds_Ratio)) + 
  geom_point(size = 4, color = "skyblue") + 
  geom_errorbar(aes(ymin = ci_lower, 
                    ymax = ci_upper), 
                width = 0.2) +  
  geom_text(aes(label = round(Odds_Ratio, 2)), 
            vjust = -0.8, 
            size = 3.5, 
            family = "Times New Roman") +
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "green") +  
  coord_flip() +  # Flip axes
  scale_x_discrete(limits = rev(sort(unique(
    race_oregon_w_WHITE$Race)))) + 
  labs(
    title = "Odds Ratio for Dismissal (Oregon, compared to White group)", 
    x = "Group", 
    y = "Odds Ratio"
  ) + 
  theme_minimal() +
  theme(
    text = element_text(family = "Times New Roman"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold")
  )

```

### corrplot() for Oregon race and exits 

```{r}
race_oregon_all_exits <- read.csv(
  "/Users/hata/Desktop/EDLD652_Diss/Data/race_oregon.csv"
)
```

```{r}
race_oregon_all_exits <- race_oregon_all_exits %>% 
  select(-2, -3)
```

### corrplot() prep 

```{r}
# making matrix for race_oregon_all_exits so I can run corrplot
race_oregon_matrix <- as.matrix(race_oregon_all_exits[, 2:7])
```

```{r}
# Run chi-squre so that I can run corrplot() 

chi_results_race_oregon <- chisq.test(race_oregon_matrix)
chi_results_race_oregon
```

Pearson's Chi-squared test — data: race_oregon_matrix X-squared = 376.19, df = 30, p-value \< 2.2e-16

```{r}
chisq.test(race_oregon_matrix, correct = FALSE)$stdres
```

```{r}
adj_std_res_race_oregon <- chisq.test(race_oregon_matrix)
```

```{r}
plot.new()
dev.off()
```

```{r}
corrplot(chi_results_race_oregon$stdres, 
         is.cor = FALSE, 
         tl.cex = 0.7)
```

```{r}
colnames(chi_results_race_oregon$stdres) <- c(
  "Withdrawal", 
  "Dismissed", 
  "Moved Out", 
  "Part B Eligible", 
  "Not Eligible", 
  "Not Determined")
```

```{r}
rownames(chi_results_race_oregon$stdres) <- c(
  "Alaska Native/American Indian", 
  "Asian", 
  "Black/African American", 
  "Hispanic/Latino", 
  "Two or More Races", 
  "Pacific Islander", 
  "White"
)
```

```{r}
chi_results_race_oregon$stdres <- chi_results_race_oregon$stdres[, order(colnames(chi_results_race_oregon$stdres))]
```

```{r}
corrplot(chi_results_race_oregon$stdres, 
         is.cor = FALSE, 
         tl.cex = 0.8, 
         tl.col = "black", 
         cl.cex = 0.7,     
         cl.offset = 1,    
         cl.ratio = 0.2)  
```
