---
title: "v2"
author: "Maiko Hata"
format: 
  pdf:
    mainfont: "Times New Roman"
    sansfont: "Times New Roman"
fig-width: 6
fig-height: 4
csl: apa.csl
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
library(rcartocolor)
library(ggplot2)
library(quarto)
```

## National and Oregon CHILD COUNTS

```{r}
# NOTE TO SELF: ADD THE CENSUS NUMBER FOR FINAL PROJECT! BIND_ROWS!! WEEK 2? 3? Labs.
```

### cc csv 1:

-   childcount1920 (all states)

-   childcount1920USOR (National and Oregon)

-   cc_transposed (childcount1920USOR but X and Y flipped)

```{r}
childcount1920 <- read_excel("/Users/maiko/Desktop/EDLD652_Diss/Data/1920-cchildcountandsettings-11.xlsx", sheet = 2)
```

```{r}
childcount1920 <- clean_names(childcount1920)
```

```{r}
childcount1920USOR <- childcount1920 %>% 
  filter(state %in% c("US and Outlying Areas", "Oregon"))
```

```{r}
childcount1920USOR <- childcount1920 %>%
  filter(state %in% c("US and Outlying Areas", "Oregon")) %>%
  select(- "race_ethnicity_total3_percent")
```

```{r}
class(childcount1920USOR)
```

```{r}
childcount1920USOR[, 2:7] <- lapply(childcount1920USOR[, 2:7], function(x) round(as.numeric(x), 2))
```

```{r}
# childcount1920 still has all the states. 

write.csv(childcount1920USOR, file = "/Users/maiko/Desktop/EDLD652_Diss/Data/childcount1920USOR.csv", row.names = FALSE)
```

```{r}
cc_transposed <- as.data.frame(t(childcount1920USOR))
cc_transposed <- rownames_to_column(cc_transposed, var = "Category")
```

```{r}
cc_transposed$Category <- c("Race/Ethnicity", 
                            "American Indian or Alaska Native", 
                            "Asian", 
                            "Black or African American", 
                            "Hispanic or Latino", 
                            "Native Hawaiian or Pacific Islander", 
                            "Two or More Races", 
                            "White")
```

```{r}
write.csv(cc_transposed, file = "/Users/maiko/Desktop/EDLD652_Diss/Data/cc_transposed.csv")
```

### cc csv 2: childcount1920USOR_long

```{r}
# making it longer yeah
childcount1920USOR_long <- childcount1920USOR %>%
 gather(key = "race", value = "percent", -state)
```

```{r}
str(childcount1920USOR_long$percent)
childcount1920USOR_long$percent <- as.numeric(childcount1920USOR_long$percent)
```

```{r}
childcount1920USOR_long <- childcount1920USOR_long %>%
  mutate(percent = round(percent, 2))
```

```{r}
write.csv(childcount1920USOR_long, file = "/Users/maiko/Desktop/EDLD652_Diss/Data/childcount1920USOR_long.csv")
```

# OLD - don't use, use the codes freom 3_chi_sq.qmd HELP: DOESN'T RUN ANYMOER! WHERE DID VAR2 go for residuals_df?

```{r}
# ggplot(residuals_df, aes(Var1, Var2, fill = Freq)) +
#  geom_tile() +
#  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
#  labs(title = "Chi-Square Standardized Residuals",
#       x = "Race", y = "Exit Reasons") +
#  theme_minimal()
```

Chi-square with Residuals: Viz 2 (CORRPLOT: https://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r#google_vignette)

# DOES IT JUST NOT WORK ON DF because it's long?

```{r}
#drop total column from race_matrix. I learned that I CANNOT USE SELECT() ON MATRIX FILES SO I CONVERTED IT BACK TO DF THEN BACK TO MATRIX AGAIN 
# race_matrix <- as.data.frame(race_matrix)
# race_matrix <- race_matrix 
# rownames(residuals_df) <- c(
#  "Alaska Native/American Indian", 
#  "Asian", 
#  "Black/African American", 
#  "Hispanic/Latino", 
#  "More than Two Races", 
#  "Pacific Islander", 
#  "White")
# already taken out 
# %>%  select(-exit_total) 
```

## Why did we make the residuals_df ???

```{r}
# residuals_df <- residuals_df[
#  order(colnames(residuals_df))
#]
```

corrplot :) Trial 1: <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

```{r}
# chisq <- chisq.test(race_matrix)
# corrplot(chisq$residuals, 
#         is.cor = FALSE) 
```

corrplot trial 2:

RENAMED THE CATEGORIES

```{r}
# dim(race_matrix)
#colnames(race_matrix)
```

# WORKING in 3_chi_dq.qmd so GO THERE, don't look here anymore NOT WORKING - help!

```{r}
# chisq <- chisq.test(race_matrix)

#Cameron 
# colnames(chisq$residuals) <- c("Withdrawal", 
#                     "Disqualification", 
#                     "Moved Out", 
#                     "Part B Eligible", 
#                     "Not Eligible", 
#                     "Not Determined")
```

REORDERED THE CATEGORIES ALPHABETICALLY I thought...?

# AND HERE, I MADE THE residual_matrix but why

```{r}
# residual_matrix <- as.matrix(chisq$residuals)
# col_order <- order(colnames(residual_matrix))
#row_order <- order(rownames(residual_matrix))
#residual_matrix <- residual_matrix[row_order, col_order]
```

```{r}
#rownames(residual_matrix) <- c(
#  "Alaska Native/American Indian" = 1, 
#  "Asian" = 2, 
#  "Black/African American" = 3, 
#  "Hispanic/Latino" = 4, 
#  "More than Two Races" = 5, 
#  "Pacific Islander" = 6, 
# "White" = 7) 
```

```{r}
# corrplot(residual_matrix, 
#         method = 'shade', 
#         is.corr = FALSE) 
```

```{r}
# maybe use this for dissertation? (FOR RESEARCHER AUDIENCE)
# corrplot(residual_matrix, 
#         method = 'number', 
#         is.corr = FALSE) 
```

```{r}
#corrplot(residual_matrix, 
#         method = 'pie', 
#         is.corr = FALSE)
```

```{r}
# USE THIS AS THE FINAL PROJECT VISUALIZATION 
#corrplot(residual_matrix, 
#         method = 'circle', 
#         is.corr = FALSE)
```
