---
title: "v2"
author: "Maiko Hata"
format: 
  pdf:
    mainfont: "Times New Roman"
    sansfont: "Times New Roman"
fig-width: 6
fig-height: 4
csl: apa.csl
execute: 
  eval: true 
  echo: false
  message: false 
  warning: false
editor: visual
engine: knitr
---

```{r}
library(tidyverse)
library(here)
library(rio)
library(knitr)
library(gt)
library(DT)
library(reactable)
library(gtsummary)
library(kableExtra)
library(tinytex)
library(janitor)
library(tidylog)
library(sjPlot)
library(lme4)
library(tibble)
library(dplyr)
library(epitools)
library(readxl)
library(pwr)
library(rcompanion)
library(grateful)
library(distill)
library(readxl)
library(scales)
library(tidyr)
library(patchwork)
library(corrplot)
library(distill)
library(tibble)
```

A. Table of 10 exit reasons

```{r Table of 10 exit reasons }
exit_categories <- import(here("Data","exit_categories.xlsx")) %>%
 clean_names() 
```

```{r Table of exit reasons original}
# ORIGINAL VERSIONS
kable(exit_categories, 
      caption = "Table of Exit Reasons",
      col.names = c('Exit Reasons','Exit Category Codes')) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

B. National and Oregon CHILD COUNTS

NOTE TO SELF: ADD THE CENSUS NUMBER FOR FINAL PROJECT! BIND_ROWS!! WEEK 2? 3? Labs.

B-1. Load data

```{r B-1 load data}
#|label: load-OSEP-CHILDCOUNT-1920-data
# Imported excel file with multiple pages but specified only the second sheet. Yay 

childcount1920 <- read_excel("/Users/hata/Desktop/EDLD652_Diss/Data/1920-cchildcountandsettings-11.xlsx", sheet = 2)
```

```{r B-1 load clean names}
#|label: load-OSEP-CHILDCOUNT-1920-clean_names()
# I cleaned names, then decided against clean_names as it messed up the column names, but it might be necessary as the file is not rendering 
childcount1920 <- clean_names(childcount1920)
```

```{r B-1 label}
#|label: load-OSEP-CHILDCOUNT-1920-filter_US_OR

childcount1920USOR <- childcount1920 %>% 
  filter(state %in% c("US and Outlying Areas", "Oregon"))
```

where did the data go wrong? Did i combine it in below? But i think I just chose and selected the one i don't need? The one below is still correct. Where is the error?

```{r B-1 delete "total 100% column"}

childcount1920USOR <- childcount1920 %>%
  filter(state %in% c("US and Outlying Areas", "Oregon")) %>%
  select(- "race_ethnicity_total3_percent")
```

B-2: chart 1:\
THIS CHUNK TO ROUND TO 2 DIGITS CONVERTED THE COLUMN TOO the distinction between the OR/US somehow. I NEED TO FIX IT

```{r B2-1 OLD codes for rounding up to 2 digits}
# YAY I PREVIOUSLY converted the character columns to NUMERIC and then reduced it to dicimals BUT IT ALSO CONVERTED state COLUMN INTO NUMERICS 
# childcount1920USOR <- childcount1920USOR %>%
#   mutate(across(where(is.character) & !any_of("state"), as.numeric, .names = "converted_{.col}")) %>%  
#   mutate(across(where(is.numeric), ~ round(.x, 2)))   

# ERROR/WARNING: 
# Warning message:
# There was 1 warning in `.fun()`.
# ℹ In argument: `across(where(is.character), as.numeric)`.
# Caused by warning:
# ! NAs introduced by coercion 
# > # Round numeric columns to 2 decimals
```

```{r B2-1 OLD}
# childcount1920USOR <- childcount1920USOR %>%
#   mutate(percent = round(percent, 2))
```

```{r B2-1 this finally worked to round to 2 digits below dicimal}
childcount1920USOR[, 2:7] <- lapply(childcount1920USOR[, 2:7], function(x) round(as.numeric(x), 2))
```

```{r B2-1 transposed to flip X and Y}
cc_transposed <- as.data.frame(t(childcount1920USOR))
cc_transposed <- rownames_to_column(cc_transposed, var = "Category")
```

```{r B2-1 kable-ing transposed dataset}
kable(cc_transposed, 
      caption = "Child Count (US \\& Oregon)") %>% 
  column_spec(1, width = "8cm") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, 
                position = "left", 
                font_size = 10.5, 
                latex_options = "scale_down")
```

B-2: visualization 2 FIX IT!!! (OLD: SOMEWHERE ALONG THE LINE I LOST THE DATA ROWS IN DF)

```{r B2-2: make longer}
# making it longer yeah
childcount1920USOR_long <- childcount1920USOR %>%
 gather(key = "race", value = "percent", -state)
```

```{r B2-2: Changed charactor to numeric so it rounds up}
str(childcount1920USOR_long$percent)
childcount1920USOR_long$percent <- as.numeric(childcount1920USOR_long$percent)
```

```{r B2-2: rounding up to 2 digits below the dicimal points}
childcount1920USOR_long <- childcount1920USOR_long %>%
  mutate(percent = round(percent, 2))
```

```{r B2-2 Chart 1 without viridis_d}
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  theme_minimal() 
```

```{r B2-2: v.1 with viridis_d}
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  theme_minimal()

```

B2 Visualization 2 v.2 - it has labels on the bar but it's ugly as hell

```{r B2-2 v.2 with viridis_d}
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity") + 
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = race), 
            position = position_stack(vjust = 0.5), 
            color = "black", 
            size = 4) +
  theme_minimal()
```

```{r B2-2 v.2 changed it to the bar chart}
ggplot(childcount1920USOR_long, aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +  # "dodge" makes it a grouped bar chart
  scale_fill_viridis_d() +  # Apply Viridis discrete color palette
  scale_y_continuous(labels = percent_format(scale = 1)) +  # Format y-axis as percentages
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  theme_minimal() 
```

```{r B2-2 v.2 changed to the bar chart AND THEN ADDED LABELS...??}
ggplot(childcount1920USOR_long, 
       aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +  
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1)) +  
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = race), 
            color = "black", 
            size = 4) +
  theme_minimal() 
```

```{r B2-2 v.2 ChatGPT suggsetion}
ggplot(childcount1920USOR_long, 
       aes(x = state, y = percent, fill = race)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.05))) +  
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "State",
       y = "Percentage",
       fill = "Race") +
  geom_text(aes(label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5,  # Moves text slightly above bars
            color = "black", 
            size = 4) +
  theme_minimal(base_size = 14) +  
  theme(legend.position = "bottom")  
```

```{r B2-2 v.2 bar chart with each of Oregon and US next to each other}
ggplot(childcount1920USOR_long, aes(x = race, y = percent, fill = state)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  # Side-by-side bars
  scale_fill_viridis_d() +  
  scale_y_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "Race",
       y = "Percentage",
       fill = "State") +
  geom_text(aes(label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5,  
            color = "black", 
            size = 4) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")  # Moves legend below for clarity

```

```{r B2-2 v.2 flip the X and Y for readability}
ggplot(childcount1920USOR_long, 
       aes(x = percent, y = race, fill = state)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +  # Side-by-side bars
  scale_fill_viridis_d() +  
  scale_x_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Percentage of Child Count by Race in US & Oregon",
       x = "Percentage",
       y = "Race",
       fill = "State") +
  geom_text(aes(label = scales::percent(percent / 100, accuracy = 0.1)), 
            position = position_dodge(width = 0.7), 
            hjust = -0.2,  
            color = "black", 
            size = 4) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")  # Moves legend below for clarity

```

C. National and Oregon EXIT data by RACE

```{r C-1 load OSEP data}
#|label: load-OSEP-EXIT-data
# importing the data and cleaning it up: 
# This is the year 2013-2022. 
# 1. REMOVED DECEASED and NOT IN USE IN OREGON CATEGORY
# 2. CREATED THE complete_or_not_eligible combining three categories 

byrace <- import(here("Data", "race v.2.xlsx")) %>% 
  clean_names() %>% 
  select(-7, -12) %>% 
  mutate(complete_or_not_eligible = complete_prior_to_reaching_max_age_for_part_c + not_eligible_for_part_b_exit_with_no_referrals + not_eligible_for_part_b_exit_with_referrals_to_other_programs)
```

I FIXED THE MISSING COLUMN by adding back part_b_eligibility_not_determined. I think this is what I can use for CHI-SQUARE WITH RESIDUALS?

```{r C-1 load OSEP data label: aggregated-by-race}

agg_by_race_and_state <- byrace %>% 
  group_by(race, area) %>% 
  summarize(exit_total = sum(exiting_total2), 
            withdrawal_by_parent = sum(withdrawal_by_parent),
            attempts_to_contact_unsuccessful = sum(attempts_to_contact_unsuccessful),
            moved_out_of_state = sum(moved_out_of_state),
            part_b_eligible_exiting_part_c = sum(part_b_eligible_exiting_part_c),
            complete_or_not_eligible = sum(complete_or_not_eligible), 
            part_b_eligibility_not_determined = sum(part_b_eligibility_not_determined))
```

I should be able to export df to excel this way but haven't tried it yet.

```{r C-1 try to export OSEP data as Excel}
# library(writexl)
# df <- tibble::tibble(A = 1:5, B = 6:10, C = 11:15)  # Example tibble
# write_xlsx(df, "my_data.xlsx")  # Saves to an Excel file 
```

agg_by_race_and_state

```{r C-1 OSEP data agg_by_race}
# aggregated by race for OR
race_oregon <- agg_by_race_and_state %>%
  filter(area == "Oregon")
```

```{r C-1 OSEP data aggregated by race for US}
race_us <- agg_by_race_and_state %>% 
  filter(area == "US and Outlying Areas") 
```

```{r C-1 OSEP data race_by_us}
race_us <- race_us %>% 
  select(-area)
```

```{r C-1 OSEP data renaming category names to full names for race_us}
race_us <- race_us %>%
  mutate(race = recode(race,
    "AM" = "Alaska Native or American Indian",
    "AS" = "Asian",
    "BL" = "Black or African American",
    "HI" = "Hispanic/Latino",
    "WH" = "White",
    "MU" = "Two or More Races",
    "PI" = "Pacific Islander"
  ))

```

```{r C-1 Made it LONGER}
us_data_long <- race_us %>% 
  pivot_longer(
    cols = 3:7, 
    values_to = "values",
    names_to = "names"
  ) %>% 
  group_by(race) %>% 
  mutate(percentage = round(values/exit_total, digits = 4))
```

```{r C-1 OSEP made it WIDER}
us_data_wide <- us_data_long %>% 
  select(-c(exit_total,values)) %>% 
  pivot_wider(
    names_from = names,
    values_from = percentage
  ) 
```

OH NO where did Part B eligibility not determined go?!?!?

```{r C-1 OSEP Changing category names in us_data_wide to clearner names}
us_data_wide <- us_data_wide %>% 
  rename(
    "Withdrawal by Parents" = withdrawal_by_parent, 
    "Disqualified" = attempts_to_contact_unsuccessful, 
    "Moved Out of State" = moved_out_of_state, 
    "Part B Eligible" = part_b_eligible_exiting_part_c, 
    "Complete or Not Eligible" = complete_or_not_eligible
  )
```

I'm trying out to see if I can do the chi-square with residuals (per <https://chatgpt.com/share/67a1833d-9fc4-8012-8193-b6fc358a9687>)

Chi-square with Residuals 1:

```{r}
race_matrix <- as.matrix(race_us[ , -1])
race_matrix <- apply(race_matrix, 2, as.numeric)
```

R doesn't like spaces or dashes / - that's why we did clean names, it could work but it can be tricky later

```{r}
rownames(race_matrix) <- c("Alaska Native/American Indian", "Asian", "Black/African American", "Hispanic/Latino", "More than Two Races", "Pacific Islander", "White")
```

Chi-square with Residuals 2:

```{r}
chi_result <- chisq.test(race_matrix)
print(chi_result)
```

Chi-square with Residuals 3:

Cameron: Residuals are what we are measuring anyway. If nothing was happening, what would be the expected values in the cells in the matrix (so the residuals = differences between expected and what we see) so it's a raw differences

```{r}
print(chi_result$stdres)
# Residual > +2 → Observed count is higher than expected.
# Residual < -2 → Observed count is lower than expected.
```

Chi-square with Residuals 4:

```{r}
residuals_df <- as.data.frame(as.table(chi_result$stdres))
```

Cameron: How can I reverse the order of Y axis (and I should delete the exit total row too)

Chi-square with Residuals: Viz 1 (HEATMAP)

```{r}
# this is race on X (columns )
ggplot(residuals_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(title = "Chi-Square Standardized Residuals",
       x = "Exit Reason", y = "Race") +
  theme_minimal() 
```

Chi-square with Residuals: Viz 2 (CORRPLOT: https://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r#google_vignette)

```{r}
colnames(race_matrix)
```

```{r}
#drop total column from race_matrix. I learned that I CANNOT USE SELECT() ON MATRIX FILES SO I CONVERTED IT BACK TO DF THEN BACK TO MATRIX AGAIN 
race_matrix <- as.data.frame(race_matrix)
race_matrix <- race_matrix %>% 
  select(-exit_total)

race_matrix <- as.matrix(race_matrix)
```

corrplot :) Trial 1: <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

```{r}
# CORRPLOT trial one
# doesn't look great at all 
# par(mar = c(5, 5, 5, 5))
# par(pin = c(6, 6))

chisq <- chisq.test(race_matrix)
corrplot(chisq$residuals, 
         is.cor = FALSE) 
```

corrplot trial 2:

```{r}
chisq <- chisq.test(race_matrix)
corrplot(chisq$residuals, 
         method = 'number', 
         is.corr = FALSE)
```

```{r}
library(ggplot2)

# Week 5 slide 13
# preview a file that would be created by ggsave()
# ggpreview <- function(...) {
#  fname <- tempfile(fileext = ".png")
#  ggsave(filename = fname, ...)
#  system2("open", fname)
#  invisible(NULL)}

# ggplot(mtcars) + 
#  aes(x = wt, y = mpg) + 
#  geom_point() 

# ggpreview(width = 3, height = 3, units = "in")
```

```{r}
# this is race on Y (rows)
# ggplot(residuals_df, aes(Var2, Var1, fill = Freq)) +  # Flip x and y
#  geom_tile() +
#  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
#  labs(title = "Chi-Square Standardized Residuals",
#       x = "Race", y = "Exit Reason") +  # Adjust labels accordingly
#  theme_minimal()

```

```{r}
# viridis_d doesn't look good, doesn't make any sense. 
# ggplot(residuals_df, aes(Var2, Var1, fill = Freq)) +  # Flip x and y
#  geom_tile() +
#  scale_fill_viridis_c(option = "magma") +  # Choose a viridis color #  labs(title = "Chi-Square Standardized Residuals",
#       x = "Race", y = "Exit Reason") +  # Adjust labels accordingly
#  theme_minimal()
```
